---
title: "The effects of severe weather events across the United States"
author: "Rocco Barbini"
date: "20 agosto 2015"
output: html_document
---

```{r setoptions, echo = FALSE, results = "hidden"}
suppressWarnings(library(knitr))
opts_chunk$set(echo = TRUE, results = "html")
options(scipen = 999, digits = 2)
```

## Synopsis

This analysis investigates the effects of the severe weather events that were recorded by NOAA across the United States between 1950 and 2011.

The data was analysed without considering the year or the State where the events took place. Moreover, since the data was nos tidy, a cleanup of the dataset was performed before the analysis. These factors should be taken into account and caution should be used when using the results as the basis for policy decisions.

The results are presented at the end of this document, separating the economic effects from those on the population health. The data processing section, which is the longer part of this report, will be of interest to the data analysts who may want to reproduce this analysis and to check if it was performed correctly.

The data set was downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 on August 20th, 2015.

## Data Processing

To process the data, I used the following libraries:
```{r, results = "hidden"}
suppressWarnings(require(dplyr))
suppressWarnings(require(ggplot2))
```

The dataset is very large and includes a variety of columns that are not relevant to my analysis. In order to make the calculations lighter on my computer, I therefore:

- loaded the data set;
- created a subset with the relevant columns;
- removed the original data set and performed an iterative garbage collection process.

I set the garbage collection function to perform three garbace collection operations spaced by one second. After some testing, i noticed that this almost always produced the best results.


```{r, cache = TRUE}
# do_collect_garbage()
#
# This function collects garbage efficiently to free up memory
#
do_collect_garbage <- function(x) {
        Sys.sleep(1)
        invisible(gc())
        Sys.sleep(1)
        invisible(gc())
        Sys.sleep(1)
        invisible(gc())
        Sys.sleep(1)
        invisible(gc())
}

storm_data <- read.csv("Data/storm_data.csv.bz2", header = T, stringsAsFactors = T)
subset <- select(storm_data, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
rm(storm_data)
do_collect_garbage()
```

Since there are too many event types, I began to clean them up by removing those that produced no damage and are therefore irrelevant to my analysis. I then cleaned up my temporary variables and freed up some more memory.

```{r, cache = TRUE}
tmp_levels <- length(levels(subset$EVTYPE))
evtypes_dmg <- aggregate(x = (subset[, "PROPDMG"] + subset[, "CROPDMG"] + subset[, "INJURIES"] + subset[, "FATALITIES"]), by = list(subset[, "EVTYPE"]), FUN = "sum")
irrelevant_evtypes <- filter(evtypes_dmg, x == 0) %>% select(Group.1)
subset <- filter(subset, !(EVTYPE %in% unlist(irrelevant_evtypes)))
subset <- droplevels(subset)
erased_levels <- tmp_levels - length(levels(subset$EVTYPE))
rm(evtypes_dmg)
rm(irrelevant_evtypes)
rm(tmp_levels)
do_collect_garbage()
```

This way I was able to remove `r erased_levels` levels with no cumulated damage to either health or property.

### Cleaning up the Event Types

I then proceeded to tidy up the EVTYPE values. This required some judgement calls, as I had to adapt the "recorded" event types to the categories quoted in NOAA documents. Those categories are the following:

```{r, cache = TRUE}
correct_event_types <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Freezing Fog", "Frost/Freeze", "Funnel Cloud", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane/Typhoon", "Ice Storm", "Lakeshore Flood", "Lake-Effect Snow", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
correct_event_types_count <- length(correct_event_types)
recorded_event_types_count <- length(levels(subset$EVTYPE))

```

Note that there are `r correct_event_types_count` admitted categories, but `r recorded_event_types_count` variations were actually recorded.

I determined the sequence of the instructions to tidy the data based on my judgement, giving priority to what looked like the most specific, then to the most important/severe events in those cases in which mixed events were reported.

Ultimately, not being an expert on this subject, I could reduce the event types only partially. In some cases, I had to invent my own categories (e.g. MUDSLIDES) in order to at least reduce the number of duplicates.

In some cases I also had to introduce a false value (e.g. COASTAL FLOOKD) so that I could use a more generic rule (e.g. to set the event type"FLOOD"), then I corrected the false value.

My first series of corrections dealt with:

- the ? value;
- waterspouts;
- tornadoes;
- thunderstorms;
- hail;
- avalanches;
- blizzards;
- dust storms;
- ice storms;
- winter storms;
- tropical storms;
- flash floods;
- coastal floods;
- lakeshore floods;
- mudslides;
- floods.


```{r, cache = TRUE}
old_event_types_count <- length(levels(subset$EVTYPE))
subset$EVTYPE <- plyr::revalue(subset$EVTYPE, c("?"="UNKNOWN"))
levels(subset$EVTYPE)<- gsub(".*waterspout.*", "WATERSPOUT", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*tornado.*", "TORNADO", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*gustnado.*", "TORNADO", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*torndao.*", "TORNADO", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("^thunderstorm hail$", "HAIL", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^thunderstorm.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*microburst.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*mircoburst.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^lightning.*thunderstorm.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^severe.*thunderstorm.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^tstm.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^tuderstorm.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^tunderstorm.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^thuderstorm.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^ tstm.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*hail.*", "HAIL", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("MARINE TSTM WIND", "MARINE THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^thun.*", "THUNDERSTORM WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*avalanche.*", "AVALANCHE", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*blizzard.*", "BLIZZARD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*dust storm.*", "DUST STORM", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*ice storm$", "ICE STORM", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*winter storm.*", "WINTER STORM", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*tropical storm.*", "TROPICAL STORM", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*flash.*flood.*", "FLASH FLOOkD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*flood.*flash.*", "FLASH FLOOkD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*coast.*flood.*", "COASTAL FLOOkD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*cstl.*flood.*", "COASTAL FLOOKD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*lake.*flood.*", "LAKESHORE FLOOkD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*mud.*", "MUDSLIDES", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*flood.*", "FLOOD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("FLASH FLOOKD", "FLASH FLOOD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("COASTAL FLOOKD", "COASTAL FLOOD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("LAKESHORE FLOOKD", "LAKESHORE FLOOD", levels(subset$EVTYPE), ignore.case = T)
subset <- droplevels(subset)
new_event_types_count <- length(levels(subset$EVTYPE))
diff_event_types_count <- old_event_types_count - new_event_types_count
rm(old_event_types_count)
do_collect_garbage()
```

This operation allowed me to clean up `r diff_event_types_count` values. But I still have `r new_event_types_count - correct_event_types_count` more values than the expected `r correct_event_types_count`.

Following the first round, I performed a second round of cleanups:

```{r, cache = TRUE}
old_event_types_count <- length(levels(subset$EVTYPE))
levels(subset$EVTYPE)<- gsub(".*devil.*", "DUST DEVIL", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("BLOWING DUST", "DUST STORM", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*drought.*", "DROUGHT", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".+heat", "EXCESSIVE HEAT", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^heat.*", "HEAT", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^(?!.*ext).*chill.*", "COLD/WINKD CHILL", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*ext.*chill.*", "EXTREME COLD/WINKD CHILL", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*lake.*snow.*", "LAKE-EFFECT SNOW", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*hea.*snow.*", "HEAVY SNOW", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*sleet.*", "SLEET", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^(?!.*marine).*high.*wind.*", "HIGH WIND", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("^(?!.*marine).*strong.*wind.*", "STRONG WIND", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*marine.*high.*wind.*", "HIGH WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*marine.*strong.*wind.*", "STRONG WIND", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^(?!thunder)(?!marine)(?!strong)(?!high).*wind.*", "WIND", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("COLD/WINKD CHILL", "COLD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("EXTREME COLD/WINKD CHILL", "EXTREME COLD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^(?!.*lake).*snow.*", "HEAVY SNOW", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*hurricane.*", "HURRICANE/TYPHOON", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*typhoon.*", "HURRICANE/TYPHOON", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*lightning.*", "LIGHTNING", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*lighting.*", "LIGHTNING", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*fire.*", "WILDFIRE", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*rain.*", "HEAVY RAIN", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*fld.*", "FLOOD", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*warm.*", "HEAT", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*coast.*storm.*", "COASTAL STORM", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*freez.*fog.*", "FREEZING FOG", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^(?!.*freez).*fog.*", "DENSE FOG", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*fre.*", "FROST/FREEZE", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*extreme.*cold.*", "EXTREME COLD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*record.*cold.*", "EXTREME COLD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*hyp.*", "EXTREME COLD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("EXTREME COLD/WIND CHILL", "EXTREME COLKD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*cold.*", "COLD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("EXTREME COLKD/WIND CHILL", "EXTREME COLD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*rip.*curr.*", "RIP CURRENT", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*wint.*we.*", "WINTER WEATHER", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*wint.*mix.*", "WINTER WEATHER", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub(".*surf.*", "HIGH SURF", levels(subset$EVTYPE), ignore.case = T)
levels(subset$EVTYPE)<- gsub("^ice.*(?<!storm)$", "ICE", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*glaze.*", "ICE", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("black ice", "ICE", levels(subset$EVTYPE), ignore.case = T, perl = T)
subset <- droplevels(subset)
new_event_types_count <- length(levels(subset$EVTYPE))
diff_event_types_count <- old_event_types_count - new_event_types_count
rm(old_event_types_count)
do_collect_garbage()
```

This second round allowed me to clean up `r diff_event_types_count` values. I still have `r new_event_types_count - correct_event_types_count` more values than the expected `r correct_event_types_count`, but the data is almost usable.

I then performed a final round of cleanups to remove most residual values:

```{r, cache = TRUE}
old_event_types_count <- length(levels(subset$EVTYPE))
levels(subset$EVTYPE)<- gsub(".*land.*", "LANDSLIDE", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("^mixed.*", "MIXED PRECIPITATION", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("HEAVY PRECIPITATION", "HEAVY RAIN", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("HEAVY MIX", "MIXED PRECIPITATION", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*storm surge.*", "STORM SURGE", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("^high$", "OTHER", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*urban.*", "OTHER", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*marine.*(?<!wind)$", "OTHER", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*erosion.*", "OTHER", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*other.*", "OTHER", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*SWELLS.*", "HEAVY SEAS", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*sea.*", "HEAVY SEAS", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*county.*", "OTHER", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*wet.*", "OTHER", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub(".*early frost.*", "FROST", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("low temperature", "COLD/WIND CHILL", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("RAPIDLY RISING WATER", "FLOOD", levels(subset$EVTYPE), ignore.case = T, perl = T)
levels(subset$EVTYPE)<- gsub("ROGUE WAVE", "TSUNAMI", levels(subset$EVTYPE), ignore.case = T, perl = T)
subset <- droplevels(subset)
new_event_types_count <- length(levels(subset$EVTYPE))
diff_event_types_count <- old_event_types_count - new_event_types_count
rm(old_event_types_count)
do_collect_garbage()
```

This time I cleaned up `r diff_event_types_count` values, which left me with `r new_event_types_count` event types. It is still `r new_event_types_count - correct_event_types_count` more values than the expected `r correct_event_types_count`, but it is a perfectly usable number.

### Cleaning up the economic data

The economic data (damage to crops and property) needs cleaning up for two reasons:

- there are different columns for the numbers and the order of magnitude;
- the order of magnitude has untidy numbers.

```{r, cache = TRUE}
levels(subset$PROPDMGEXP)
levels(subset$CROPDMGEXP)
```

I assumed that H means hundred, K means thousand, M means million and B means billion. I disregarded all other values.

```{r, cache = TRUE}
levels(subset$PROPDMGEXP)<- gsub("h", "H", levels(subset$PROPDMGEXP))
levels(subset$PROPDMGEXP)<- gsub("k", "K", levels(subset$PROPDMGEXP))
levels(subset$PROPDMGEXP)<- gsub("m", "M", levels(subset$PROPDMGEXP))
levels(subset$PROPDMGEXP)<- gsub("B", "B", levels(subset$PROPDMGEXP))
levels(subset$CROPDMGEXP)<- gsub("h", "H", levels(subset$CROPDMGEXP))
levels(subset$CROPDMGEXP)<- gsub("k", "K", levels(subset$CROPDMGEXP))
levels(subset$CROPDMGEXP)<- gsub("m", "M", levels(subset$CROPDMGEXP))
levels(subset$CROPDMGEXP)<- gsub("B", "B", levels(subset$CROPDMGEXP))

subset[subset$PROPDMGEXP == "H", "PROPDMG"] = subset[subset$PROPDMGEXP == "H", "PROPDMG"] * 100
subset[subset$PROPDMGEXP == "K", "PROPDMG"] = subset[subset$PROPDMGEXP == "K", "PROPDMG"] * 1000
subset[subset$PROPDMGEXP == "M", "PROPDMG"] = subset[subset$PROPDMGEXP == "M", "PROPDMG"] * 1000000
subset[subset$PROPDMGEXP == "B", "PROPDMG"] = subset[subset$PROPDMGEXP == "B", "PROPDMG"] * 1000000000

subset[subset$PROPDMGEXP == "H", "CROPDMG"] = subset[subset$PROPDMGEXP == "H", "CROPDMG"] * 100
subset[subset$PROPDMGEXP == "K", "CROPDMG"] = subset[subset$PROPDMGEXP == "K", "CROPDMG"] * 1000
subset[subset$PROPDMGEXP == "M", "CROPDMG"] = subset[subset$PROPDMGEXP == "M", "CROPDMG"] * 1000000
subset[subset$PROPDMGEXP == "B", "CROPDMG"] = subset[subset$PROPDMGEXP == "B", "CROPDMG"] * 1000000000
```

After this transformation, I removed some more columns to make te data lighter.

```{r, cache = TRUE}
subset <- select(subset, EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG)
do_collect_garbage()
```

### Disclaimer: judgement calls

Not being an expert on the subject matter of this analysis, I had to rely on many judgement calls when I tidied up the data. Such judgement calls included the decision to create some new categories, which were not present in the official NOAA documentation, and deciding what to aggregate as OTHER.

The quality of this research can be greatly improved with the help of an expert in meteorology.

### Analysis: Effects on the population health

The 10 event types that produced the most cumulated fatalities are the following:

```{r, cache = TRUE}
pop_health <- select(subset, EVTYPE, FATALITIES, INJURIES)
pop_health <- group_by(pop_health, EVTYPE) %>% summarize(fatalities = sum(FATALITIES), injuries = sum(INJURIES))
res_pop_health_topsumfatal <- head(arrange(pop_health, desc(fatalities)), 10)
res_pop_health_topsumfatal
```

The 10 event types that produced the most cumulated injuries are the following:

```{r, cache = TRUE}
res_pop_health_topsuminj <- head(arrange(pop_health, desc(injuries)), 10)
res_pop_health_topsuminj
```

The previous tables are heavily influenced by the frequency of the events. If we average the data to see the effects of the individual events in each category, the following plot shows how the event types are distributed based on the mean injuries and fatalities of each.

```{r, cache = TRUE}
pop_health_mean <- select(subset, EVTYPE, FATALITIES, INJURIES)
pop_health_mean <- group_by(pop_health_mean, EVTYPE) %>% summarize(fatalities = mean(FATALITIES), injuries = mean(INJURIES))

inj_fat_plot_mean <- ggplot(data = pop_health_mean, aes(x = fatalities, y = injuries)) 
inj_fat_plot_mean <- inj_fat_plot_mean + geom_point(shape = 1, size = 2, col = "red")
inj_fat_plot_mean <- inj_fat_plot_mean + scale_x_continuous(name = "Fatalities")
inj_fat_plot_mean <- inj_fat_plot_mean + scale_y_continuous(name = "Injuries")
inj_fat_plot_mean <- inj_fat_plot_mean + ggtitle("Mean injuries and Fatalities across all event types")
inj_fat_plot_mean <- inj_fat_plot_mean + theme(plot.title = element_text(lineheight=1.5, face="bold", size = 16))
inj_fat_plot_mean
```

As we can see from the plot, most events produce on average less than 2 injuries and less than 1 fatality.
The worst events, i.e. those that produce on average more than 2 injuries and/or 1 fatality, are the following:

```{r, cache = TRUE}
res_pop_health_moredamage <- filter(pop_health_mean, injuries >= 2 | fatalities >= 1)
res_pop_health_moredamage
```

The absolute worst events, i.e. those that produce on average more than 2 injuries and 1 fatality, are the following:

```{r, cache = TRUE}
res_pop_health_mostdamage <- filter(pop_health_mean, injuries >= 2 & fatalities >= 1)
res_pop_health_mostdamage
```

In general, the 10 event types that produced the most mean fatalities per single event are the following:

```{r, cache = TRUE}
pop_health <- select(subset, EVTYPE, FATALITIES, INJURIES)
pop_health <- group_by(pop_health, EVTYPE) %>% summarize(fatalities = mean(FATALITIES), injuries = mean(INJURIES))
res_pop_health_morefatal <- head(arrange(pop_health, desc(fatalities)), 10)
res_pop_health_morefatal
```

The 10 event types that produced the most mean injuries per single event are the following:

```{r, cache = TRUE}
res_pop_health_moreinjuries <- head(arrange(pop_health, desc(injuries)), 10)
res_pop_health_moreinjuries
```

The 10 most numerous events are the following (with averaged outcome):

```{r, cache = TRUE}
pop_health_count <- select(subset, EVTYPE, FATALITIES, INJURIES)
pop_health_count <- cbind(pop_health_count, count = rep(1, nrow(pop_health_count)))
pop_health_count <- group_by(pop_health_count, EVTYPE) %>% summarize(number_of_events = sum(count), fatalities = mean(FATALITIES), injuries = mean(INJURIES))
head(arrange(pop_health_count, desc(number_of_events)), 10)

```

Finally, the 10 most numerous events that produced on averate at least 1 casualty or 1 fatality are the following:

```{r, cache = TRUE}
top_10_worst_pop <- filter(pop_health_count, injuries >= 1 | fatalities >= 1) %>% arrange(desc(number_of_events))
res_pop_health_worstpop <- head(top_10_worst_pop, 10)
res_pop_health_worstpop
```

### Analysis: Effects on the economy

I can find the event types that produced the most cumulated economic damages by aggregating per event type the sum of the damage to property and crops:

```{r}
eco_dmg <- select(subset, EVTYPE, PROPDMG, CROPDMG)
eco_dmg_total <- group_by(eco_dmg, EVTYPE) %>% summarize(total_economic_loss = sum(PROPDMG) + sum(CROPDMG))
res_eco_dmg_topdmg <- head(arrange(eco_dmg_total, desc(total_economic_loss)), 10)
res_eco_dmg_topdmg
```

The event types that produced the most cumulated damage to property are the following:
```{r}
eco_dmg_prop <- group_by(eco_dmg, EVTYPE) %>% summarize(total_economic_loss = sum(PROPDMG))
res_eco_dmg_topprop <- head(arrange(eco_dmg_prop, desc(total_economic_loss)), 10)
res_eco_dmg_topprop
```

The event types that produced the most cumulated damage to property are the following:
```{r}
eco_dmg_crop <- group_by(eco_dmg, EVTYPE) %>% summarize(total_economic_loss = sum(CROPDMG))
res_eco_dmg_topcrop <- head(arrange(eco_dmg_crop, desc(total_economic_loss)), 10)
res_eco_dmg_topcrop
```

If we consider the mean damages per single event, these are the ones that produced the worst damages on property:

```{r}
eco_dmg_mean <- group_by(eco_dmg, EVTYPE) %>% summarize(property_dmg = mean(PROPDMG), crop_dmg = mean(CROPDMG))

res_eco_moreprop <- head(arrange(eco_dmg_mean, desc(property_dmg)), 10)
res_eco_moreprop
```

These, on the other hand, are the ones that produced the worst damages on crops:

```{r}
res_eco_morecrop <- head(arrange(eco_dmg_mean, desc(crop_dmg)), 10)
res_eco_morecrop
```

It is also interesting to see if property damage and crop damage are correlated (to make the plot readable, I removed the two outliers that emerged from the previous analyses: HURRICANE/TYPHOON and FLOOD):

```{r}
prop_crop_plot_mean <- ggplot(data = eco_dmg_mean[which(eco_dmg_mean$property_dmg < 100000000),], aes(x = property_dmg, y = crop_dmg)) 
prop_crop_plot_mean <- prop_crop_plot_mean + geom_point(shape = 1, size = 2, col = "red")
prop_crop_plot_mean <- prop_crop_plot_mean + scale_x_continuous(name = "Property Damage")
prop_crop_plot_mean <- prop_crop_plot_mean + scale_y_continuous(name = "Crop Damage")
prop_crop_plot_mean <- prop_crop_plot_mean + ggtitle("Mean property and crop damage across all event types")
prop_crop_plot_mean <- prop_crop_plot_mean + theme(plot.title = element_text(lineheight=1.5, face="bold", size = 16))
prop_crop_plot_mean
```

Compared to the injuries vs. fatalities plot, most event types produce significant damage either to property or to crops, with only a handful procuring significant damages to both. 

The worst events, i.e. those producing on average at least than 1 million dollars of damage to either crops or property per single event, are the following:

```{r, cache = TRUE}
res_eco_moredamage <- filter(eco_dmg_mean, property_dmg >= 1000000 | crop_dmg >= 1000000)
res_eco_moredamage
```

The absolute worst, those producing on average at least than 1 million dollars of damage to both crops and property per single event, are the following:

```{r, cache = TRUE}
res_eco_mostdamage <- filter(eco_dmg_mean, property_dmg >= 1000000 & crop_dmg >= 1000000)
res_eco_mostdamage
```

The 10 most numerous events are the following (with averaged outcome):

```{r, cache = TRUE}
prop_crop_count <- select(subset, EVTYPE, PROPDMG, CROPDMG)
prop_crop_count <- cbind(prop_crop_count, count = rep(1, nrow(prop_crop_count)))
prop_crop_count <- group_by(prop_crop_count, EVTYPE) %>% summarize(number_of_events = sum(count), property_dmg = mean(PROPDMG), crop_dmg = mean(CROPDMG))
head(arrange(prop_crop_count, desc(number_of_events)), 10)
```

Finally, the most numerous events that produced on averate at least 1 casualty or 1 fatality are the following:

```{r, cache = TRUE}
res_eco_worsteco <- filter(prop_crop_count, property_dmg >= 1000000 | crop_dmg >= 1000000) %>% arrange(desc(number_of_events))
res_eco_worsteco <- head(res_eco_worsteco, 10)
res_eco_worsteco
```

## Results

Risk is usually defined as the product of the probability and severity of an event. Paraphrasing this definition, to define the most harmful types of events it is necessary to consider:

- how frequent they are;
- how severe they are.

In this analysis, we define **severity** as *the magnitude of the mean effects per single even*. We also define **frequency** as *the number of times that an event was recorded*.

An event that is very frequent, but of little consequence, does not constitute a real risk. On the other hand, a severe event that almost never happens is less of a risk than a less severe events that happened many times.

### Across the United States, which types of events are most harmful with respect to population health?

With respect to population health, the most severe events, i.e. those that produce on average more than 2 injuries and/or 1 fatality per single event, are the following:

```{r}
res_pop_health_moredamage
```

The absolute worst events, i.e. those that produce on average more than 2 injuries and 1 fatality, are the following:

```{r, cache = TRUE}
res_pop_health_mostdamage
```

The 10 event types that produced the most mean fatalities per single event are the following:

```{r}
res_pop_health_morefatal
```

The 10 event types that produced the most mean injuries per single event are the following:

```{r}
res_pop_health_moreinjuries
```

Finally, the most numerous events that produced on averate at least 1 casualty or 1 fatality are the following:

```{r}
res_pop_health_worstpop
```

### Across the United States, which types of events have the greatest economic consequences?

With respect to the economy, the most severe events, i.e. those that produce on average more than 1 million dollars of property and/or crop damage per single event, are the following:

```{r}
res_eco_moredamage
```

The absolute worst events, i.e. those that produce on average more than 1 million dollars of property and crop damage, are the following:

```{r, cache = TRUE}
res_eco_mostdamage
```

The 10 event types that produced the most mean property damages per single event are the following:

```{r}
res_eco_moreprop
```

The 10 event types that produced the most mean crop damages per single event are the following:

```{r}
res_eco_morecrop
```

Finally, the most numerous events that produced on averate at least 1 casualty or 1 fatality are the following:

```{r}
res_eco_worsteco
```